Fractional Types

### 2013-02-17 frac.tex
beginning of sec3

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no notion of \emph{input} or \emph{output}; more
precisely, because of strict preservation of information, these two concepts
coincide.  This is the root of reversibility.

But if we want to model functions of any flavor, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position to possess negative information, as it
is really a request for information.  Since information is logarithmic, this
means that a request for information should behave like a \emph{fractional
  type}.

\roshan{It is unclear how fractions help differentiate between inputs
  and outputs. On first reading this seemed promising, but nothing
  in the rest of the paper substantiates this point of view.}

\roshan{I think we should arrive at relational programming as a means
  of giving semantics to fractionals, rather than the other way
  round. To show that relational programming is reversible, one does
  not need to resort to fractionals. Everyone understand that
  relations can be reversed and none of this setup with {{Pi}} is
  required for that. Our key point should be to communicate that
  fractionals have a clean semantics in the setting of relational
  programming rather than the other way round.}

@@

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
we think of a value like {{1/left ()}} as a \emph{first-class constraint}
that can only be satisfied if it is matched with an actual value {{left ()}}
or as representing the absence of some information (i.e., \emph{negative
  information}), that can only be supplied by {{left ()}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} represents a fission point which creates --- out
of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.  The best analogy is probably that of vectors and co-vectors,
also know as column vectors and row vectors (respectively): in the finite
dimensional case, they are isomorphic, but they are nevertheless quite
different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
since we are explicitly interested in \emph{isomorphisms}, we should not
conflate negative-information values with positive-information values just
because they happen to be isomorphic.

In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain the
simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
in this section, once we make the extension to higher-order functions, we
must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

@@

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below.

%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[ b ]]  
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

@@

%%%%%%%%%%%%%
\subsection{Relational Semantics}

@@

%%%%%%%%%%%%%%%%
\subsection{Evaluation}

@@

%%%%%%%%%%%%%%%%%
%% \subsection{Model}

%%%%%%%%%%%%%%%%%
\subsection{Completeness}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semantics}

@@

\amr{stuck}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expressiveness}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Examples}

@@

>>>>

The language {{Pi}} models isomorphisms of values rather well.  However,
purposefully, it has no distinguished notion of \emph{input} or
\emph{output}; more precisely, because of strict preservation of information,
these two concepts coincide.  This is the root of reversibility. But if we
want to model functions of any flavor as values, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position to possess negative information, as it
is really a request for information.  Since information is logarithmic, this
means that a request for information should behave like a \emph{fractional
  type}.

% AMR: a function a -o b is modeled by (1/a,b) not (a,b) 
% \roshan{It is unclear how fractions help differentiate between inputs
%   and outputs. On first reading this seemed promising, but nothing
%   in the rest of the paper substantiates this point of view.}
%

% AMR: relations are indeed only presented as a way to explain the semantics
% \roshan{I think we should arrive at relational programming as a means
%   of giving semantics to fractionals, rather than the other way
%   round. To show that relational programming is reversible, one does
%   not need to resort to fractionals. Everyone understand that
%   relations can be reversed and none of this setup with {{Pi}} is
%   required for that. Our key point should be to communicate that
%   fractionals have a clean semantics in the setting of relational
%   programming rather than the other way round.}

@@

\begin{definition}[Syntax of {{PiTF}}]
\label{def:PiT}
We extend Def.~\ref{def:Pi} with:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
%
% base iso ::= ... | eta*_b | eps*_b
where:
{{eta*_b : 1 <-> 1/b * b : eps*_b}}
\end{definition}

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.  However,
we think of a value like {{1/left ()}} as a \emph{first-class constraint}
that can only be satisfied if it is matched with an actual value {{left ()}}
or as representing the absence of some information (i.e., \emph{negative
  information}), that can only be supplied by {{left ()}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} represents a fission point which creates --- out
of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.  The best analogy is probably that of vectors and co-vectors,
also know as column vectors and row vectors (respectively): in the finite
dimensional case, they are isomorphic, but they are nevertheless quite
different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words,
since we are explicitly interested in \emph{isomorphisms}, we should not
conflate negative-information values with positive-information values just
because they happen to be isomorphic. (See Sec.~\ref{sec:cat} for a more
thorough discussion about the meaning of fractional types and values in an
algebraic and categorical sense.)

For now, we proceed with an operational understanding of the language. The
semantics we provide is simple: each combinator {{c : b1 <-> b2}} is
interpreted as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively.

\begin{definition}[Relational semantics]
\label{def:relational-PiTF}
Each combinator {{c : b1 <-> b2}} in {{PiTF}} denotes a relation
as specified below. 
%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%----
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%----
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%----
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[ b ]]  
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()
\end{definition}

\newpage
~
\newpage

In the previous section, we have been able to maintain the simple idea that a
syntactic isomorphism denotes a \emph{bijection}. In richer languages, e.g.,
in a Turing complete language that admits non-termination, isomorphisms
necessarily denote \emph{partial
  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
in this section, once we make the extension to higher-order functions, we
must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

@@

@@

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expressiveness and Examples}

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:cat}
